{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-03 paper.tex
sc pass

The outline of this paper is the following. In
Sect.~\ref{sec:FastKernel} we present the technical details of the
implementation of the {\tt FastKernel} method. Then, in
Sect.~\ref{sec:benchmark} we illustrate timings and performance
results of the {\tt APFELgrid} library and grids. Finally, in
Sect.~\ref{sec:conclusion} we conclude and discuss possible future
developments.

@@

\subsection{Features and limitations of {\tt FK} tables}
The {\tt FK} product of Eqn.~\ref{eq:FK} differs with respect to the product in Eqn.~\ref{eq:applconv} in several notable ways. Firstly 
the typical {\tt APPLgrid} or {\tt FastNLO} products require PDFs at a general scale $Q^2$, which implies performing PDF evolution e.g Eqn.~\ref{eq:fastPDFfinal_recalled} for every variation of the PDFs at the input scale during the fit. However in the {\tt FK} convolution this evolution
is pre-cached at the stage of {\tt FK} table generation, requiring only initial-scale PDFs as an input at fitting time. This pre-caching of the evolution
also removes the need for performing the sums over hard scale and perturbative order during the fit, further reducing the number of operations required.
The {\tt FK} convolution also benefits from the typically reduced number of active partonic modes at the fitting scale, with the sum over flavours in Eqn.~\ref{eq:FK} being limited to those directly used in the fit. Having reduced the calculation to such a simple form, it is also possible to straightforwardly
apply standard and widely-available computational tools such as multi-threading through e.g OpenMP or Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to further reduce computational cost.

While these features provide significant performance enhancements, the {\tt FK } table format is not suitable as a complete replacement for tools
such as {\tt APPLgrid}. The pre-computation of the PDF evolution necessarily means that all theory parameters such as perturbative order, strong coupling and factorization/renormalization scales are inextricably embedded in each {\tt FK} table. In order to perform PDF fits including variations of these parameters, multiple {\tt FK } tables must be computed, each with different theory settings. While performing such a re-calculation directly from Monte Carlo codes
would be exceptionally time consuming, the data representation in {\tt APPLgrid} files allows for an efficient combination with varying theory parameters.

@@

While sheer computational speed is typically the primary consideration when computing observables
in a PDF fit, other factors such as table size in the filesystem and memory, along with the computational cost of
pre-computing {\tt FK} tables must be considered. Indeed, the computation of the {\tt FK} table in Eqn.~\ref{eq:FKTable}
requires a great deal of operations which can be time consuming, particuarly in the case of source {\tt APPLgrids} with very high
interpolation precision.

In Fig.~\ref{fig:performance} we examine the {\tt FK} table generation time with {\tt APFELgrid}, {\tt FK} table file size and memory usage of the {\tt FK} tables arising from the same source {\tt APPLgrids} as discussed in Fig.~\ref{fig:timings}. 
When examining the table generation time per point, we observe timings from a few milliseconds to
3.5 minutes per point, with differences arising from the varying grid densities used in the input {\tt APPLgrid} files. 
In terms of the grid size on disc, {\tt FK} tables are typically larger than their corresponding {\tt APPLgrids}, primarily as the {\tt FK} file format is encoding in plain text for compatibility whereas {\tt APPLgrids} are expressed in binary as {\tt ROOT} files. However when
measuring the in-memory resident set size used by the both convolution procedures, the amount of memory required by {\tt FK} tables is systematically less than {\tt APPLgrid} for all processes considered here by at least 75\%.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering                                      
  \includegraphics[scale=0.6]{plots/t0b}
\caption{\small {\tt APFELgrid} generation time per point and
  comparison to {\tt APPLgrid} for grid size and resident set size
  (RSS).}
\label{fig:performance}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We have shown that in several practical examples the numerical evaluation of an {\tt FK}
product is much faster than the corresponding {\tt APPLgrid} product, even in the case where
neither SIMD or multi-threading are applied. {\tt FK} tables are supplied in a simple plain-text 
format in order to simplify the construction of user interfaces, and therefore typically are often larger 
than the corresponding {\tt APFELgrid} file. However we have shown that the in-memory resident set size occupied by {\tt
  FK} tables is typically smaller than that requred by {\tt APPLgrid} by at least 75\%.

>>>>

The outline of this paper is the following. In
Sect.~\ref{sec:FastKernel} we present the technical details of the
implementation of the {\tt FastKernel} method. Then, in
Sect.~\ref{sec:benchmark} we illustrate timings and performance
results of the {\tt APFELgrid} library and grids. Finally, in
Sect.~\ref{sec:conclusion} we summarise the results discussed in this
work.

@@

\subsection{Features and limitations of {\tt FK} tables}

The {\tt FK} product of Eq.~(\ref{eq:FK}) differs with respect to the
product in Eq.~(\ref{eq:applconv}) in several notable ways. Firstly
the typical {\tt APPLgrid} or {\tt FastNLO} products require PDFs at a
general scale $Q^2$, which implies performing PDF evolution {\it e.g.}
Eq.~(\ref{eq:fastPDFfinal_recalled}) for every variation of the PDFs
at the input scale during the fit. However in the {\tt FK} convolution
this evolution is pre-cached at the stage of {\tt FK} table
generation, requiring only initial-scale PDFs as an input at fitting
time. This pre-caching of the evolution also removes the need for
performing the sums over hard scale and perturbative order during the
fit, further reducing the number of operations required.  The {\tt FK}
convolution also benefits from the typically reduced number of active
partonic modes at the fitting scale, with the sum over flavours in
Eq.~(\ref{eq:FK}) being limited to those directly used in the
fit. Having reduced the calculation to such a simple form, it is also
possible to straightforwardly apply standard and widely-available
computational tools such as multi-threading through e.g OpenMP or
Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to
further reduce computational cost.

While these features provide significant performance enhancements, the
{\tt FK} table format is not suitable as a complete replacement for
tools such as {\tt APPLgrid}. The pre-computation of the PDF evolution
necessarily means that all theory parameters such as perturbative
order, strong coupling and factorization/renormalization scales are
inextricably embedded in each {\tt FK} table. In order to perform PDF
fits including variations of these parameters, multiple {\tt FK}
tables must be computed, each with different theory settings. While
performing such a re-calculation directly from Monte Carlo codes would
be exceptionally time consuming, the data representation in {\tt
  APPLgrid} files allows for an efficient combination with varying
theory parameters.

@@

While sheer computational speed is typically the primary consideration
when computing observables in a PDF fit, other factors such as table
size in the filesystem and memory, along with the computational cost
of pre-computing {\tt FK} tables must be considered. Indeed, the
computation of the {\tt FK} table in Eq.~(\ref{eq:FKTable}) requires a
great deal of operations which can be time consuming, particuarly in
the case of source {\tt APPLgrids} with very high interpolation
precision.

In Fig.~\ref{fig:performance} we examine the {\tt FK} table generation
time with {\tt APFELgrid}, {\tt FK} table file size and memory usage
of the {\tt FK} tables arising from the same source {\tt APPLgrids} as
discussed in Fig.~\ref{fig:timings}.  When examining the table
generation time per point, we observe timings from a few milliseconds
to 3.5 minutes per point, with differences arising from the varying
grid densities used in the input {\tt APPLgrid} files.  In terms of
the grid size on disk, {\tt FK} tables are typically larger than their
corresponding {\tt APPLgrids}, primarily as the {\tt FK} file format
is encoding in plain text for compatibility whereas {\tt APPLgrids}
are expressed in binary as {\tt ROOT} files. However when measuring
the in-memory resident set size used by the both convolution
procedures, the amount of memory required by {\tt FK} tables is
systematically less than {\tt APPLgrid} for all processes considered
here by at least 75\%.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering                                      
  \includegraphics[scale=0.6]{plots/t0b}
\caption{\small {\tt APFELgrid} generation time per point and
  comparison to {\tt APPLgrid} for grid size on disk and resident set size
  (RSS).}
\label{fig:performance}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We have shown that in several practical examples the numerical
evaluation of an {\tt FK} product is much faster than the
corresponding {\tt APPLgrid} product, even in the case where neither
SIMD or multi-threading are applied. {\tt FK} tables are supplied in a
simple plain-text format in order to simplify the construction of user
interfaces, and therefore typically are often larger than the
corresponding {\tt APPLgrid} file. However we have shown that the
in-memory resident set size occupied by {\tt FK} tables is typically
smaller than that required by {\tt APPLgrid} by at least 75\%.
